KiCS2: A New Compiler from Curry to Haskell

### 2011-05-03 papers/wflp2011/paper.tex
Paper update w.r.t. reviewers' comments

\addtolength{\textheight}{0.2cm}
\addtolength{\topmargin}{-0.1cm}
\addtolength{\textwidth}{0.2cm}
\addtolength{\oddsidemargin}{-0.1cm}
\addtolength{\evensidemargin}{-0.1cm}

@@

\newcommand{\code}[1]{\mbox{\small\texttt{#1}}}
\newcommand{\ccode}[1]{``\code{#1}''}
\newcommand{\bs}{\char92\xspace} % backslash
\newcommand{\us}{\char95\xspace} % underscore
\newcommand{\dollar}{\code{\char36}} % dollar in programs
\newcommand{\Choice}[2]{#1 \mathop{?} #2}

% ------------------------------------------------------------------
% SYMBOLS & FORMULAS

\newcommand{\funset}{\ensuremath{_{\cal S}}}

\begin{document}
\pagestyle{plain}
\sloppy

@@

\begin{abstract}
In this paper we present our first steps towards a new system
to compile functional logic programs of the source language Curry
into purely functional Haskell programs.
Our implementation is based on the idea to represent
non-deterministic values in the respective data structure.
This enables the application of various search strategies
to extract values from the search space.
We show by several benchmarks that our implementation
can compete with or outperforms other existing implementations of Curry.
\end{abstract}

@@

\begin{figure}[t]
\[
\begin{array}{@{}rcl@{~~~~~~~~}l}
  e & ::= & x & \mbox{$x$ is a variable}\\
    &  |  & c(e_1,\ldots,e_n) & \mbox{$c$ is an $n$-ary constructor symbol}\\
    &  |  & f(e_1,\ldots,e_n) & \mbox{$f$ is an $n$-ary function symbol}\\
    &  |  & \Choice{e_1}{e_2} & \mbox{choice} \\[1ex]
%    &  |  & \mbox{\code{let}}~\{x_1=e_1;\ldots;x_n = e_n\}~\code{in}~e
%                                            & \mbox{let binding}\\[1ex]
  D & ::= & f(x_1,\ldots,x_n) = e
                & \mbox{$n$-ary function $f$ with a single rule}\\
    &  |  & f(c(y_1,\ldots,y_m),x_2,\ldots,x_n) = e
                & \mbox{matching rule for $n$-ary function $f$}\\
    &     & & \mbox{$c$ is an $m$-ary constructor symbol}\\
  P & ::= & D_1 \ldots D_k
\end{array}
\]
\caption{Uniform programs ($e$: expressions, $D$: definitions, $P$: programs)}
\label{fig:uniform}
\end{figure}
%
Finally, we assume that the pattern matching strategy
is explicitly encoded in individual matching functions.
In contrast to \cite{AlbertHanusHuchOliverVidal05},
where the pattern matching strategy is encoded in case expressions,
we assume that each case expression is transformed into
a new operation in order to avoid complications by translating
nested case expressions.
Thus, we assume that all programs are \emph{uniform}
according to the definition in Fig.~\ref{fig:uniform}.
There, the variables in the left-hand sides of each rule are pairwise
different, and
the constructors in the left-hand sides of the matching rules of each function
are pairwise different.
Uniform programs have a simple form of pattern matching:
either a function is defined by a single rule without pattern matching,
or it is defined by rules with only one constructor in the left-hand
side of each rule, and in the same argument for all rules.%
\footnote{For simplicity, we require in
Fig.~\ref{fig:uniform} that the matching argument is always the
first one, but one can also choose any other argument.}
For instance, the operation \code{xor} defined above can be transformed
into the following uniform program:
%
\begin{curry}
  xor True   x = xor' x
  xor False  x = x
  xor' False = True
  xor' True  = False
\end{curry}
%
In particular, there are no overlapping rules for functions
(except for the choice operation \ccode{?} which is considered as predefined).
Antoy \cite{Antoy01PPDP} showed that each functional logic program,
i.e., each constructor-based conditional term rewriting system,
can be translated into an equivalent unconditional term rewriting system
without overlapping rules but containing choices in the right-hand sides,
also called LOIS (limited overlapping inductively sequential) system.
Furthermore, Brassel \cite{Brassel11Thesis} showed the semantical
equivalence of narrowing computations in LOIS systems
and rewriting computations in uniform programs.
Due to these results, uniform programs are a reasonable intermediate
language for our translation into Haskell which will be presented
in the following sections.

@@

  xor' False          = True
  xor' True           = False
  xor' (Choice x1 x2) = Choice (xor' x1) (xor' x2)
\end{curry}
%
The operation \code{xorSelf} is not defined by a pattern matching rule
and, thus, need not be changed.
If we evaluate the expression \ccode{xorSelf aBool}, we get the result
\begin{lstlisting}
  Choice (Choice False True) (Choice True False)
\end{lstlisting}
How can we interpret this result?
In principle, the choices represent different possible values.
Thus, if we want to show the different values of an expression
(which is usually the task of a top-level ``read-eval-print'' loop),
we enumerate all values contained in the choices.
These are \code{False}, \code{True}, \code{True}, and \code{False}
in the result above.
Unfortunately, this is not conform with the call-time choice semantics
discussed in Section~\ref{ctc-semantics} which should not yield
a value like \code{True}.
The call-time choice semantics requires that the choice
of a value made for
the initial expression \code{aBool} should be consistent
in the entire computation.
For instance, if we select the value \code{False} for the
expression \code{aBool}, this selection should be made
at all other places where this expression might have been copied
during the computation. However, our initial implementation
duplicates the initially single \code{Choice} into finally three
occurrences of \code{Choice}.

@@

To summarize, our implementation is based on the following principles:
\begin{enumerate}
\item Each non-deterministic choice is represented by a \code{Choice}
constructor with a unique identifier.
\item When matching a \code{Choice} constructor, the choice is
moved to the result of this operation with the same identifier,
i.e., a non-deterministic argument
yields non-deterministic results for each of the argument's values.
\item Each choice occurring in a computation gets its own unique
identifier.
\end{enumerate}
The latter principle requires the creation of fresh identifiers
during a computation---a non-trivial problem in functional languages.
One possibility is the use of a global counter
that is accessed by unsafe features whenever a new identifier
is required. Unfortunately, unsafe features inhibit the use
of optimization techniques developed for purely functional programs
and make the application of advanced evaluation and search
strategies (e.g., parallel strategies) more complex.
Therefore, we avoid unsafe features in our implementation.
Instead, we thread some global information
through our program in order to supply fresh references
at any point of a computation.
For this purpose, we assume a type \code{IDSupply}
with operations
\begin{haskell}
  thisID      :: IDSupply -> ID
  leftSupply  :: IDSupply -> IDSupply
  rightSupply :: IDSupply -> IDSupply
\end{haskell}
and add a new argument of type \code{IDSupply} to each operation
of the source program.

\code{thisID} creates an identifier from a current
identifier supply, and
\code{leftSupply} and \code{rightSupply} create
new different identifier supplies (where both do not supply the ID obtained
by \code{thisID}) from a given one.
The latter are used when an operation calls two\footnote{The
extension to more than two is straightforward.}
other operations in the right-hand side of a rule. In this case,
the other operations must be supplied with their individual distinct
identifier supplies. For instance, the operation \code{main}
defined by
\label{sec:xor-main}
\begin{curry}
  main = xorSelf aBool
\end{curry}
is translated into
\begin{haskell}
  main s = xorSelf (aBool (leftSupply s)) (rightSupply s)
\end{haskell}
Any choice in the right-hand side of a rule gets its own identifier
by the operation \code{thisID}, as in
\begin{haskell}
  aBool s = Choice (thisID s) True False
\end{haskell}
There are various implementations of \code{IDSupply}.
The simplest implementation uses unbounded integers:
\begin{haskell}
  type IDSupply = Integer
  leftSupply  n = 2*n
  rightSupply n = 2*n+1
  thisID      n = n
\end{haskell}
There are other more sophisticated implementations available
\cite{AugustssonRittriSynek94}.
Actually, our compilation system is parameterized over different
implementations of \code{IDSupply}
in order to perform some experiments and choose
the most appropiate for a given application.
Each implementation must ensure that, if $s$ is a value of type
\code{IDSupply}, then \code{thisID($o_1$($\ldots$($o_n$ $s$)$\ldots$))}
and \code{thisID($o'_1$($\ldots$($o'_m$ $s$)$\ldots$))}
are different identifiers provided that
$o_i,o'_j \in \{\code{leftSupply},\code{rightSupply}\}$
and $o_1\cdots o_n \neq o'_1\cdots o'_m$.

@@

So far, our generated operations compute all the non-deterministic
values of an expression represented by a structure containing
\code{Choice} constructors. In order to extract the various
values from this structure, we have to define operations
that compute all possible choices in some order where the choice identifiers
are taken into account.
To provide a common interface for such operations, we introduce
a data type to represent the general outcome of a computation,
\begin{haskell}
  data Try a = Val a | Choice ID a a | Fail
\end{haskell}
together with an auxiliary operation:\footnote{Note that the
operation \code{try} is not really polymorphic but overloaded
for each data type and, therefore, defined in instances of some type class.}
\begin{haskell}
  try :: a -> Try a
  try (Choice i x y) = Choice i x y
  try Fail           = Fail
  try x              = Val x
\end{haskell}
In order to take the identity of choices into account when extracting values,
one has to remember which choice (e.g., left or right branch)
has been made for some particular choice.
Therefore, we introduce the type
\begin{haskell}
  data Choice = NoChoice | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoChoice} represents the fact that a choice has not yet been made.
Furthermore, we need operations to lookup the current choice
for a given identifier or change its choice:
\begin{haskell}
  lookupChoice :: ID -> IO Choice
  setChoice    :: ID -> Choice -> IO ()
\end{haskell}
In Haskell, there are different possibilities to implement a mapping
from choice identifiers to some value of type \code{Choice}.
Our implementation support various options together with different
implementations of \code{IDSupply}.
For instance, a simple implementation can be obtained by
using updatable values, i.e., the Haskell type \code{IORef}.
In this case, choice identifiers are memory cells instead of integers:\footnote{%
Consequently, integers cannot be used for \code{IDSupply}s. Instead,
\code{IDSupply}s can be implemented as an infinite tree containing
all memory cells required in a computation that are created
on demand with initial value \code{NoChoice}.}
\begin{haskell}
  newtype ID = ID (IORef Choice)
\end{haskell}
The implementation of the lookup and set operations is straightforward:
\begin{haskell}
  lookupChoice (ID ref) = readIORef ref
  setChoice (ID ref) c = writeIORef ref c
\end{haskell}
Now we can print all values contained in a choice structure
in a depth-first manner by the following operation:
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: Try a -> IO ()
  printValsDFS (Val v)          = print v
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupChoice i >>= choose
   where
    choose ChooseLeft  = printValsDFS (try x1)
    choose ChooseRight = printValsDFS (try x2)
    choose NoChoice    = do newChoice ChooseLeft  x1
                            newChoice ChooseRight x2

@@

For instance, to print all values of the expression \code{main}
defined in Section~\ref{sec:xor-main},
we evaluate the Haskell expression \ccode{printValsDFS (try (main inits))}
where \code{inits} is the initial value for \code{IDSupply}
(where all references are initialized with \code{NoChoice}).
Thus, we obtain the output
\begin{haskell}
  False
  False
\end{haskell}
In general, one has to propagate all choices to the top level
of a computation before printing the results.
This can be obtained by computing the normal form of the main
expression before passing it to the operation \code{try}.
Thus, the result values of \code{main} can be
printed by evaluating
\ccode{printValsDFS (try (id $\dollar$!! main inits))}
where \ccode{$f$ \$!! $x$} %$
denotes the application of the operation $f$ to the normal form
of its argument $x$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Benchmarks}
\label{sec:Benchmarks}
% Benchmarks from April 11 on climens

@@

\begin{figure}
\centering
\begin{tabular}{|@{~~}l@{~~}|*{5}{@{~~}r@{~~}|}}
\hline
System  & ReverseHO & Primes & PrimesPeano & Queens & QueensUser \\\hline
KiCS2   &      oom  &   1.22 &       0.30  &  10.02 &     13.08  \\
KiCS2HO &     0.24  &   0.09 &       0.27  &   0.65 &      0.73  \\
PAKCS   &     7.97  &  14.52 &      23.08  &  81.72 &     81.98  \\
MCC     &     0.27  &   0.32 &       1.77  &   3.25 &      3.62  \\
MonC    &      oom  &  16.74 &        oom  &    oom &       oom  \\
GHC     &     0.24  &   0.06 &       0.22  &   0.06 &      0.11  \\
\hline
\end{tabular}
\caption{Benchmarks: higher-order functional programs}
 \label{fig:bench-higher-order}
\end{figure}
%
The next collection of benchmarks (Fig.~\ref{fig:bench-higher-order})
considers higher-order functional programs so that we drop
the comparison to first-order Prolog systems.
``ReverseHO'' reverses a list with one million
elements in linear time using higher-order functions
like \code{foldl} and \code{flip}.
``Primes'' computes the the 2000th prime number via the
sieve of Eratosthenes using higher-order functions,
and ``PrimesPeano'' computes the 256th prime number but
with Peano numbers and user-defined lists.
Finally, ``Queens'' (and ``QueensUser'' with user-defined lists)
computes the number of safe positions of $11$ queens on a $11 \times 11$
chess board.

@@

In this section we sketch some additional features of our implementation.
Due to lack of space, we cannot discuss all details of them.

@@

\begin{figure}
\centering
\begin{tabular}{|@{~~}l@{~~}|*{3}{@{~~}r@{~~}|}}
\hline
Method               & PermSort & PermSortPeano &  NDNums \\\hline
\code{printValsDFS}  &    2.82  &         3.66  & $\infty$ \\
depth-first search   &    5.33  &         6.09  & $\infty$ \\
breadth-first search &   26.16  &        29.25  &    34.00 \\
iterative deepening  &    9.16  &        10.91  &     0.16 \\
%parallel (1 proc.)   &    5.46  &         6.29  & $\infty$ \\
%parallel (2 proc.)   &    3.16  &         3.49  & $\infty$ \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different search strategies}
 \label{fig:search-strategies}
\end{figure}
%
In order to compare the various search strategies,
Fig.~\ref{fig:search-strategies} contains some corresponding
benchmarks.
``PermSort'' and ``PermSortPeano'' are the programs
discussed in Fig.~\ref{fig:bench-nondet} and ``NDNums'' is the program
\begin{curry}
  f n = f (n+1) ? n
\end{curry}
where we look for the first solution of \ccode{f 0 == 25000}
(obviously, depth-first search strategies do not terminate on this
equation).
All strategies except for the ``direct print'' method \code{printValsDFS}
translate choice structures into monadic list structures
in order to print them according to the user options.
The benchmarks show that the overhead of this transformation
is acceptable so that this more flexible approach is the default one.

@@

% TODO: set functions??

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{sec:Conclusions}

@@

Due to the representation of non-deterministic values as data,
our system easily supports various search strategies
in constrast to Prolog-based implementations.
Since we compile Curry programs into Haskell,
we can exploit the implementation efforts done
for functional programming.
Hence, purely functional parts of functional logic programs
can be executed with almost the same efficiency as Haskell programs.
Our benchmarks show that even the execution of the
non-deterministic parts can compete with other
implementations of Curry.
Furthermore, we can also exploit advanced developments
in the implementation of Haskell, like the parallel evaluation of
expressions, which is an interesting topic for future work.
Finally, our implementation has many opportunities
for optimization, like
better program analyses to approximate purely deterministic computations.

\bibliographystyle{plain}
\bibliography{mh}

>>>>

@@

\newcommand{\code}[1]{\mbox{\small\texttt{#1}}}
\newcommand{\ccode}[1]{``\code{#1}''}
\newcommand{\bs}{\char92\xspace} % backslash
\newcommand{\us}{\char95\xspace} % underscore
\newcommand{\dollar}{\code{\char36}} % dollar in programs
\newcommand{\Choice}[2]{#1 \mathop{?} #2}
\newcommand{\funset}{\ensuremath{_{\cal S}}}

\begin{document}
%\pagestyle{plain}
\sloppy

@@

\begin{abstract}
In this paper we present our first steps towards a new system
to compile functional logic programs of the source language Curry
into purely functional Haskell programs.
Our implementation is based on the idea to represent
non-deterministic results as values in the data types corresponding
to the results.
This enables the application of various search strategies
to extract values from the search space.
We show by several benchmarks that our implementation
can compete with or outperforms other existing implementations of Curry.
\end{abstract}

@@

\begin{figure}[t]
\[
\begin{array}{@{}rcl@{~~~~~~~~}l}
  e & ::= & x & \mbox{$x$ is a variable}\\
    &  |  & c(e_1,\ldots,e_n) & \mbox{$c$ is an $n$-ary constructor symbol}\\
    &  |  & f(e_1,\ldots,e_n) & \mbox{$f$ is an $n$-ary function symbol}\\
    &  |  & \Choice{e_1}{e_2} & \mbox{choice} \\[1ex]
  D & ::= & f(x_1,\ldots,x_n) = e
                & \mbox{$n$-ary function $f$ with a single rule}\\
    &  |  & f(c(y_1,\ldots,y_m),x_2,\ldots,x_n) = e
                & \mbox{matching rule for $n$-ary function $f$}\\
    &     & & \mbox{$c$ is an $m$-ary constructor symbol}\\
  P & ::= & D_1 \ldots D_k
\end{array}
\]
\caption{Uniform programs ($e$: expressions, $D$: definitions, $P$: programs)}
\label{fig:uniform}
\end{figure}
%
Finally, we assume that the pattern matching strategy
is explicitly encoded in individual matching functions.
In contrast to \cite{AlbertHanusHuchOliverVidal05},
where the pattern matching strategy is encoded in case expressions,
we assume that each case expression is transformed into
a new operation in order to avoid complications by translating
nested case expressions.
Thus, we assume that all programs are \emph{uniform}
according to the definition in Fig.~\ref{fig:uniform}.\footnote{%
Uniform programs have been also considered in
\cite{MorenoEtAl90ALP} to define lazy narrowing strategies
for functional logic programs. Although the motivations are similar,
our notion of uniform program is more restrictive since we
allow only a single non-variable argument in each left-hand side
of a rule.}
There, the variables in the left-hand sides of each rule are pairwise
different, and
the constructors in the left-hand sides of the matching rules of each function
are pairwise different.
Uniform programs have a simple form of pattern matching:
either a function is defined by a single rule without pattern matching,
or it is defined by rules with only one constructor in the left-hand
side of each rule, and in the same argument for all rules.%
\footnote{For simplicity, we require in
Fig.~\ref{fig:uniform} that the matching argument is always the
first one, but one can also choose any other argument.}
For instance, the operation \code{xor} defined above can be transformed
into the following uniform program:
%
\begin{curry}
  xor True   x = xor' x
  xor False  x = x
  xor' False = True
  xor' True  = False
\end{curry}
%
In particular, there are no overlapping rules for functions
(except for the choice operation \ccode{?} which is considered as predefined).
Antoy \cite{Antoy01PPDP} showed that each functional logic program,
i.e., each constructor-based conditional term rewriting system,
can be translated into an equivalent unconditional term rewriting system
without overlapping rules but containing choices in the right-hand sides,
also called LOIS (limited overlapping inductively sequential) system.
Furthermore, Brassel \cite{Brassel11Thesis} showed the semantical
equivalence of narrowing computations in LOIS systems
and rewriting computations in uniform programs.
Due to these results, uniform programs are a reasonable intermediate
language for our translation into Haskell which will be presented
in the following sections.

@@

  xor' False          = True
  xor' True           = False
  xor' (Choice x1 x2) = Choice (xor' x1) (xor' x2)
\end{curry}
%
The operation \code{xorSelf} is not defined by a pattern matching rule
and, thus, need not be changed.
If we evaluate the expression \ccode{xorSelf aBool}, we get the result
\begin{lstlisting}
  Choice (Choice False True) (Choice True False)
\end{lstlisting}
How can we interpret this result?
In principle, the choices represent different possible values.
Thus, if we want to show the different values of an expression
(which is usually the task of a top-level ``read-eval-print'' loop),
we enumerate all values contained in the choices.
These are \code{False}, \code{True}, \code{True}, and \code{False}
in the result above.
Unfortunately, this does not conform to the call-time choice semantics
discussed in Section~\ref{ctc-semantics} which should not yield
a value like \code{True}.
The call-time choice semantics requires that the choice
of a value made for
the initial expression \code{aBool} should be consistent
in the entire computation.
For instance, if we select the value \code{False} for the
expression \code{aBool}, this selection should be made
at all other places where this expression might have been copied
during the computation. However, our initial implementation
duplicates the initially single \code{Choice} into finally three
occurrences of \code{Choice}.

@@

To summarize, our implementation is based on the following principles:
\begin{enumerate}
\item Each non-deterministic choice is represented by a \code{Choice}
constructor with a unique identifier.
\item When matching a \code{Choice} constructor, the choice is
moved to the result of this operation with the same identifier,
i.e., a non-deterministic argument
yields non-deterministic results for each of the argument's values.
\item Each choice introduced in a computation gets its own unique
identifier.
\end{enumerate}
The latter principle requires the creation of fresh identifiers
during a computation---a non-trivial problem in functional languages.
One possibility is the use of a global counter
that is accessed by unsafe features whenever a new identifier
is required. Unfortunately, unsafe features inhibit the use
of optimization techniques developed for purely functional programs
and make the application of advanced evaluation and search
strategies (e.g., parallel strategies) more complex.
Therefore, we avoid unsafe features in our implementation.
Instead, we thread some global information
through our program in order to supply fresh references
at any point of a computation.
For this purpose, we assume a type \code{IDSupply}
with operations
\begin{haskell}
  initSupply  :: IO IDSupply
  thisID      :: IDSupply -> ID
  leftSupply  :: IDSupply -> IDSupply
  rightSupply :: IDSupply -> IDSupply
\end{haskell}
and add a new argument of type \code{IDSupply} to each operation
of the source program, i.e., a Curry operation of type
\begin{curry}
  f :: $\tau_1$ ->$~\cdots~$->$~\tau_n~$->$~\tau$
\end{curry}
is translated into a Haskell function of type
\begin{haskell}
  f :: $\tau_1$ ->$~\cdots~$->$~\tau_n~$-> IDSupply ->$~\tau$
\end{haskell}
%
Conceptually, one can consider \code{IDSupply}
as an infinite set of identifiers that is created at the beginning
of an evaluation by the operation \code{initSupply}.
The operation \code{thisID} takes some identifier from this set, and
\code{leftSupply} and \code{rightSupply} splits this set
into two disjoint subsets without the identifier
obtained by \code{thisID}.
The split operations \code{leftSupply} and \code{rightSupply}
are used when an operation calls two\footnote{The
extension to more than two is straightforward.}
other operations in the right-hand side of a rule. In this case,
the other operations must be supplied with their individual disjoint
identifier supplies. For instance, the operation \code{main}
defined by
\label{sec:xor-main}
\begin{curry}
  main :: Bool
  main = xorSelf aBool
\end{curry}
is translated into
\begin{haskell}
  main :: IDSupply -> Bool
  main s = xorSelf (aBool (leftSupply s)) (rightSupply s)
\end{haskell}
Any choice in the right-hand side of a rule gets its own identifier
by the operation \code{thisID}, as in
\begin{haskell}
  aBool s = Choice (thisID s) True False
\end{haskell}
There are various implementations of \code{IDSupply}.
The simplest implementation uses unbounded integers:
\begin{haskell}
  type IDSupply = Integer
  initSupply    = return 1
  thisID      n = n
  leftSupply  n = 2*n
  rightSupply n = 2*n+1
\end{haskell}
There are other more sophisticated implementations available
\cite{AugustssonRittriSynek94}.
Actually, our compilation system is parameterized over different
implementations of \code{IDSupply}
in order to perform some experiments and choose
the most appropiate for a given application.
Each implementation must ensure that, if $s$ is a value of type
\code{IDSupply}, then \code{thisID($o_1$($\ldots$($o_n$ $s$)$\ldots$))}
and \code{thisID($o'_1$($\ldots$($o'_m$ $s$)$\ldots$))}
are different identifiers provided that
$o_i,o'_j \in \{\code{leftSupply},\code{rightSupply}\}$
and $o_1\cdots o_n \neq o'_1\cdots o'_m$.

@@

So far, our generated operations compute all the non-deterministic
values of an expression represented by a structure containing
\code{Choice} constructors. In order to extract the various
values from this structure, we have to define operations
that compute all possible choices in some order where the choice identifiers
are taken into account.
To provide a common interface for such operations, we introduce
a data type to represent the general outcome of a computation,
\begin{haskell}
  data Try a = Val a | Choice ID a a | Fail
\end{haskell}
together with an auxiliary operation:\footnote{Note that the
operation \code{try} is not really polymorphic but overloaded
for each data type and, therefore, defined in instances of some type class.}
\begin{haskell}
  try :: a -> Try a
  try (Choice i x y) = Choice i x y
  try Fail           = Fail
  try x              = Val x
\end{haskell}
In order to take the identity of choices into account when extracting values,
one has to remember which choice (e.g., left or right branch)
has been made for some particular choice.
Therefore, we introduce the type
\begin{haskell}
  data Choice = NoChoice | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoChoice} represents the fact that a choice has not yet been made.
Furthermore, we need operations to lookup the current choice
for a given identifier or change its choice:
\begin{haskell}
  lookupChoice :: ID -> IO Choice
  setChoice    :: ID -> Choice -> IO ()
\end{haskell}
In Haskell, there are different possibilities to implement a mapping
from choice identifiers to some value of type \code{Choice}.
Our implementation supports various options together with different
implementations of \code{IDSupply}.
For instance, a simple but efficient implementation can be obtained by
using updatable values, i.e., the Haskell type \code{IORef}.
In this case, choice identifiers are memory cells instead of integers:
\begin{haskell}
  newtype ID = ID (IORef Choice)
\end{haskell}
Consequently, an implementation of \code{IDSupply} requires
an infinite set of memory cells which can represented as a tree structure:
\begin{haskell}
  data IDSupply = IDSupply ID IDSupply IDSupply
  thisID      (IDSupply r _ _) = r
  leftSupply  (IDSupply _ s _) = s
  rightSupply (IDSupply _ _ s) = s
  \end{haskell}
The infinite tree of memory cells (with initial value \code{NoChoice})
can be constructed as follows,
where the operation \code{unsafeInterleaveIO} is used to construct
the tree on demand:
\begin{haskell}
  initSupply = getIDTree
  getIDTree = do s1 <- unsafeInterleaveIO getIDTree
                 s2 <- unsafeInterleaveIO getIDTree
                 r  <- unsafeInterleaveIO (newIORef NoChoice)
                 return (IDSupply (ID r) s1 s2)
  \end{haskell}
Using memory cells,
the implementation of the lookup and set operations is straightforward:
\begin{haskell}
  lookupChoice (ID ref) = readIORef ref
  setChoice (ID ref) c = writeIORef ref c
\end{haskell}
Now we can print all values contained in a choice structure
in a depth-first manner by the following operation:
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: Try a -> IO ()
  printValsDFS (Val v)          = print v
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupChoice i >>= choose
   where
    choose ChooseLeft  = printValsDFS (try x1)
    choose ChooseRight = printValsDFS (try x2)
    choose NoChoice    = do newChoice ChooseLeft  x1
                            newChoice ChooseRight x2

@@

For instance, to print all values of the expression \code{main}
defined in Section~\ref{sec:xor-main},
we evaluate the Haskell expression
\begin{haskell}
  initSupply >>= \s -> printValsDFS (try (main s))
\end{haskell}
Thus, we obtain the output
\begin{haskell}
  False
  False
\end{haskell}
In general, one has to propagate all choices and failures to the top level
of a computation before printing the results.
Otherwise, the operation \code{try} applied to an expression
like \ccode{Just aBool} would return a \code{Val}-structure
instead of a \code{Choice}
so that the main operation \code{printValsDFS} would miss
the non-determinism of the result value.
Therefore, we have to compute the normal form of the main
expression before passing it to the operation \code{try}.
Hence, the result values of \code{main} are printed by evaluating
\begin{haskell}
  initSupply >>= \s -> printValsDFS (try (id $\dollar$!! main s))
\end{haskell}
where \ccode{$f$ \$!! $x$} %$
denotes the application of the operation $f$ to the normal form
of its argument $x$.
This has the effect that a choice or failure occurring somewhere in
a computation will be moved (by the operation \ccode{\$!!})
to the root of the main expression
so that the corresponding search strategy can process it.
This ensures that, after the computation to a normal form,
an expression without a \code{Choice} or \code{Fail} at the root
is a value, i.e., it does not contain a \code{Choice} or \code{Fail}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Benchmarks}
\label{sec:Benchmarks}

@@

\begin{figure}
\centering
\begin{tabular}{|@{~~}l@{~~}|*{5}{@{~~}r@{~~}|}}
\hline
System  & ReverseHO & Primes & PrimesPeano & Queens & QueensUser \\\hline
KiCS2   &      oom  &   1.22 &       0.30  &  10.02 &     13.08  \\
KiCS2HO &     0.24  &   0.09 &       0.27  &   0.65 &      0.73  \\
PAKCS   &     7.97  &  14.52 &      23.08  &  81.72 &     81.98  \\
MCC     &     0.27  &   0.32 &       1.77  &   3.25 &      3.62  \\
MonC    &      oom  &  16.74 &        oom  &    oom &       oom  \\
GHC     &     0.24  &   0.06 &       0.22  &   0.06 &      0.11  \\
\hline
\end{tabular}
\caption{Benchmarks: higher-order functional programs}
 \label{fig:bench-higher-order}
\end{figure}
%
The next collection of benchmarks (Fig.~\ref{fig:bench-higher-order})
considers higher-order functional programs so that we drop
the comparison to first-order Prolog systems.
``ReverseHO'' reverses a list with one million
elements in linear time using higher-order functions
like \code{foldl} and \code{flip}.
``Primes'' computes the 2000th prime number via the
sieve of Eratosthenes using higher-order functions,
and ``PrimesPeano'' computes the 256th prime number but
with Peano numbers and user-defined lists.
Finally, ``Queens'' (and ``QueensUser'' with user-defined lists)
computes the number of safe positions of $11$ queens on a $11 \times 11$
chess board.

@@

In this section we sketch some additional features of our implementation.
Due to lack of space, we cannot discuss them in detail.

@@

\begin{figure}
\centering
\begin{tabular}{|@{~~}l@{~~}|*{3}{@{~~}r@{~~}|}}
\hline
Method               & PermSort & PermSortPeano &  NDNums \\\hline
\code{printValsDFS}  &    2.82  &         3.66  & $\infty$ \\
depth-first search   &    5.33  &         6.09  & $\infty$ \\
breadth-first search &   26.16  &        29.25  &    34.00 \\
iterative deepening  &    9.16  &        10.91  &     0.16 \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different search strategies}
 \label{fig:search-strategies}
\end{figure}
%
In order to compare the various search strategies,
Fig.~\ref{fig:search-strategies} contains some corresponding
benchmarks.
``PermSort'' and ``PermSortPeano'' are the programs
discussed in Fig.~\ref{fig:bench-nondet} and ``NDNums'' is the program
\begin{curry}
  f n = f (n+1) ? n
\end{curry}
where we look for the first solution of \ccode{f 0 == 25000}
(obviously, depth-first search strategies do not terminate on this
equation).
All strategies except for the ``direct print'' method \code{printValsDFS}
translate choice structures into monadic list structures
in order to print them according to the user options.
The benchmarks show that the overhead of this transformation
is acceptable so that this more flexible approach is the default one.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions and Related Work}
\label{sec:Conclusions}

@@

Due to the representation of non-deterministic values as data,
our system easily supports various search strategies
in constrast to Prolog-based implementations.
Since we compile Curry programs into Haskell,
we can exploit the implementation efforts done
for functional programming.
Hence, purely functional parts of functional logic programs
can be executed with almost the same efficiency as Haskell programs.
Our benchmarks show that even the execution of the
non-deterministic parts can compete with other
implementations of Curry.

We have already discussed in the introduction the various
efforts to implement functional logic languages,
like the construction of abstract machines
\cite{AntoyHanusLiuTolmach05,HanusSadre99JFLP,Lux99FLOPS} and
the compilation into Prolog \cite{AntoyHanus00FROCOS}
or Haskell \cite{BrasselFischerHanusReck11,BrasselHuch07,BrasselHuch09}.
Our benchmarks show that an efficient implementation
by compiling into
a functional language depends on carefully handling
the sharing of non-deterministic choices. For instance, our previous
implementation \cite{BrasselFischerHanusReck11}, where sharing
is explicitly managed by the monadic techniques proposed
in \cite{FischerKiselyovShan09}, has not satisfied the expectations
that came from the benchmarks reported in \cite{FischerKiselyovShan09}.
Due to these experiences, we used in our new compiler the
compilation scheme initially proposed in \cite{BrasselFischer08IFL}
which produces much faster code, as shown in our benchmarks.

If non-deterministic results are collected in data structures,
one has more fine-grained control over non-deterministic steps.
For instance, \cite{AlqaddoumiAntoyFischerReck10} proposes
pull-tab steps to move non-determinism from arguments to
the result position of a function. Antoy \cite{Antoy11ICLP}
shows that single pull-tab steps are semantics-preserving.
Thus, it is not necessary to move each choice to the root
of an expression, as done in our implementation,
but one could also perform further local computations in the arguments
of a choice before moving it up. This might be a reasonable strategy
if all non-deterministic values are required but many computations fail.
However, the general effects of such refinements need further
investigations.

Our implementation has many opportunities for optimization, like
better program analyses to approximate purely deterministic computations.
We can also exploit advanced developments
in the implementation of Haskell, like the parallel evaluation of
expressions. These are interesting topics for future work.

\begin{thebibliography}{10}

\bibitem{AlbertHanusHuchOliverVidal05}
E.~Albert, M.~Hanus, F.~Huch, J.~Oliver, and G.~Vidal.
\newblock Operational semantics for declarative multi-paradigm languages.
\newblock {\em Journal of Symbolic Computation}, 40(1):795--829, 2005.

\bibitem{AlqaddoumiAntoyFischerReck10}
A.~Alqaddoumi, S.~Antoy, S.~Fischer, and F.~Reck.
\newblock The pull-tab transformation.
\newblock In {\em Proc. of the Third International Workshop on Graph
  Computation Models}, pages 127--132. Enschede, The Netherlands, 2010.
\newblock Available at
  http://gcm-events.org/gcm2010/pages/gcm2010-preproceedings.pdf.

\bibitem{Antoy01PPDP}
S.~Antoy.
\newblock Constructor-based conditional narrowing.
\newblock In {\em Proc.\ of the 3rd International ACM SIGPLAN Conference on
  Principles and Practice of Declarative Programming (PPDP 2001)}, pages
  199--206. ACM Press, 2001.

\bibitem{Antoy11ICLP}
S.~Antoy.
\newblock On the correctness of the pull-tab transformation.
\newblock In {\em To appear in Proceedings of the 27th International Conference
  on Logic Programming (ICLP 2011)}, 2011.
